DEG_RNA_summary

### 2025-02-04 05b_Functional_enrichment_analysis/DEG_gProfiler.Rmd
Cleaned up code

@@

###Delete below ###########

RNAlistRDS005 %>% dplyr::filter(logFC > 0.5) %>% nrow()
RNAlistRDS005 %>% dplyr::filter(logFC < -0.5) %>% nrow()

RNAlistRDS005 <- RNAlistRDS005 %>% dplyr::filter(annotation_reference == "Tan")
###plotting
plot(density(RNAlistRDS005$AveExpr))
# bar plot of DEGs by condition
df <- RNAlistRDS005 %>%
  dplyr::group_by(cluster_id, stage, clustering,annotation_reference) %>%
  dplyr::summarise(DEGs = n(), median_logFC = median(logFC), median_AveExpr = median(AveExpr), median_padjloc = median(p_adj.loc)) %>%
  ungroup()
ggplot(df, aes(x=stage, y = DEGs, fill=cluster_id)) +
  geom_bar(stat='identity') +
  scale_y_continuous(expand = c(0,0)) +
  RotatedAxis() +
  xlab('Sample ID') +
  facet_wrap(~clustering,scales = "free_x", ncol = 4)
ggsave(paste0(dir_out, "Barplot_cluster_id_stage_clustering.pdf"),
  plot = last_plot(),
  width = 15, height = 6)
sum(df$DEGs)
# bar plot of DEGs by condition with AveExpr > 18.5
#There are extremely many DEGs in the clusters eF, epithelial and C7_Activated_Fibro, too many to be reliable. Median AveExpr seems to be lower in those clusters and gProfiler gave good results with a threshold of AveExpr>18.5. Therefore, I set the cut-off at AveExpr>18.5
df18_3 <- RNAlistRDS005 %>%
  dplyr::filter(AveExpr > 18.3) %>%
  dplyr::group_by(cluster_id, stage, clustering,annotation_reference) %>%
  dplyr::summarise(DEGs = n(), median_logFC = median(logFC), median_AveExpr = median(AveExpr), median_padjloc = median(p_adj.loc)) %>%
  ungroup()
df18_5 <- RNAlistRDS005 %>%
  dplyr::filter(AveExpr > 18.5) %>%
  dplyr::group_by(cluster_id, stage, clustering,annotation_reference) %>%
  dplyr::summarise(DEGs = n(), median_logFC = median(logFC), median_AveExpr = median(AveExpr), median_padjloc = median(p_adj.loc)) %>%
  ungroup()
ggplot(df18_5, aes(x=stage, y = DEGs, fill=cluster_id)) +
  geom_bar(stat='identity') +
  scale_y_continuous(expand = c(0,0)) +
  RotatedAxis() +
  xlab('Sample ID') +
  facet_wrap(~clustering,scales = "free_x", ncol = 4) +
  ggtitle("AveExpr > 18.5")
ggsave(paste0(dir_out, "Barplot_cluster_id_stage_clustering_AveExpr185.pdf"),
  plot = last_plot(),
  width = 15, height = 6)
ggplot(df18_5, aes(x=stage, y = DEGs, fill=cluster_id)) +
  geom_bar(stat='identity') +
  scale_y_continuous(expand = c(0,0)) +
  RotatedAxis() +
  xlab('Sample ID')+
  ggtitle("AveExpr > 18.5") 
ggsave(paste0(dir_out, "Barplot_cluster_id_stage_AveExpr185.pdf"),
  plot = last_plot(),
  width = 6, height = 4)
sum(df18_5$DEGs)

# Count the number of genes in each combination of clustering and stage
RNAcounts <- table(RNAlistRDS005$clustering, RNAlistRDS005$stage)
pheatmap(RNAcounts)
# Create the grouped bar plot
ggplot(as.data.frame(RNAcounts), aes(x = Var1, y = Freq, fill = Var2)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Clustering", y = "Count", fill = "Stage") +
  ggtitle("Count of Genes by Clustering and Stage") +
  theme_bw() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Create vulcano plot
ggplot(RNAlistRDS005, aes(x = logFC, y = -log10(p_adj.loc), color = cluster_id)) +
  geom_point(data = RNAlistRDS005) +
  theme_minimal() +
  xlim(-0.8,1.3)
ggsave(paste0(dir_out, "Vulcano_logFC_padjloc_clusterid.pdf"),
  plot = last_plot(),
  width = 6, height = 4)

#scatter plot logFC, AveExpr
ggplot(RNAlistRDS005, aes(x = AveExpr, y = logFC, color = cluster_id)) +
  geom_point(data = RNAlistRDS005) +
  theme_minimal()
ggsave(paste0(dir_out, "Scatter_logFC_AveExpr_clusterid.pdf"),
  plot = last_plot(),
  width = 6, height = 4)
plot(density(log2(RNAlistRDS005$AveExpr-16.97529)))
```

@@

>>>>

For Shaoline, also looking at BH readjustment
```{r}
##Combine muscat output into one dataframe
PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/A030"), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
PATHS <- c("/home/common/data/output/projects/ENDO/E044/A030/major_celltypes_Tan/data/res_mm_dream.rds",
           "/home/common/data/output/projects/ENDO/E044/A030/myeloid_lymphocyte_epithelial_endothelial_Tan/data/res_mm_dream.rds",
           "/home/common/data/output/projects/ENDO/E044/A030/stromal_Tan/data/res_mm_dream.rds",
           "/home/common/data/output/projects/ENDO/E044/A030/united_celltypes_Tan/data/res_mm_dream.rds")

# Load data from files
print(PATHS)
RNAlistRDS <- lapply(PATHS, readRDS)

# Assign names based on PATHS
names(RNAlistRDS) <- PATHS

#combine sub-lists
RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE)

# Filter without BH adjustment 
##BH adjustment before AveExpr filtering is identical to p_adj.loc, BH adjustment after AveExpr filtering leads to very few significant genes
RNAlistRDS005 <- lapply(RNAlistRDS, function(x) {
  dplyr::filter(x, AveExpr > 18.53202 & p_adj.loc < 0.05)
})
RNAlistRDS005 <- as.data.frame(do.call(rbind, RNAlistRDS005))

##exclude ribosomal genes (comment: does not change the inflection point)
rb.genes_DEGs <- unique(grep(pattern = "^RP[SL]", RNAlistRDS005$gene, value=TRUE)) #around 150 unique rb genes
RNAlistRDS005 <- RNAlistRDS005[!(RNAlistRDS005$gene %in% rb.genes_DEGs), ]

# Filter with BH adjustment after AveExpr filtering leads to very few significant genes
RNAlistRDS005BH <- lapply(RNAlistRDS, function(x) {
  x <- dplyr::filter(x, AveExpr > 18.53202)
  x$p_adj.BH <- p.adjust(x$p_val, method = "BH")
  dplyr::filter(x, p_adj.BH < 0.05)
})
RNAlistRDS005BH <- as.data.frame(do.call(rbind, RNAlistRDS005BH))

@@

@@

##Ongoing xxxx
gProfiler based on RNA assay and per annotation with gene expression >18.53202 and BH pval adjusted
```{r}
##function to run gProfiler and save output
gProfiler_FUN <- function(x,ANALYSIS,GRADE,ORDERED_QUERY,ASSAY,dir_out) {
#gProfiler
custom_bg <- unique(x$gene)
query <- x %>%
  dplyr::filter(AveExpr > 18.53202) %>%
  dplyr::filter(p_adj.loc < 0.05) %>%
  dplyr::arrange(p_adj.loc) #arrange by lowest p_adj.loc
if(nrow(query) > 0){
query <- distinct(query, gene, .keep_all= TRUE) #remove duplicates of gene with higher p_adj.loc
query <- query$gene
gostres <<- gost(query = query, 
                organism = "hsapiens", 
                ordered_query = ORDERED_QUERY, #can use TRUE, p_adj.loc has been arranged 
                multi_query = FALSE, 
                significant = TRUE, #only statistically significant results should be returned
                exclude_iea = FALSE, 
                measure_underrepresentation = FALSE, 
                evcodes = TRUE,
                user_threshold = 0.05, 
                correction_method = "g_SCS", #g_SCS is default
                domain_scope = "annotated", #if custom_bg is set, I believe this shouldn't matter too much
                custom_bg = custom_bg, #custom background, use!
                numeric_ns = "", 
                sources = NULL, #by default all sources are analyzed
                as_short_link = FALSE)
if(length(gostres) > 0){
results <- gostres$result
results$cluster_id <- levels(as.factor(x$cluster_id))
results$analysis <- print(ANALYSIS)
results$assay <- print(ASSAY)
results$grade <- print(GRADE) 
results$ordered_query <- print(ORDERED_QUERY)
results$parents <- NULL
write.csv(results,paste0("gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".csv"))
print(paste0("Saved gotres as gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".csv"))
plot <- gostplot(gostres, capped = TRUE, interactive = FALSE) + ggtitle(paste0("gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY)))
ggsave(plot = plot, filename = paste0(dir_out,"gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".png"), width = 7, height = 5)
print("Saved plot")
}
}
}
##function to run gProfiler_FUN on different grades, analysis and assay
RUN_gProfier <- function(GRADE, ANALYSIS,ASSAY,SPECIFICATION){
dir_out <- paste0("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044",ANALYSIS,"/",ASSAY,"assay_",SPECIFICATION,"/")
dir.create(dir_out, recursive = TRUE)
setwd(dir_out)
#Read paths to DEG output .rds files
if (ASSAY == "RNA") { 
    PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/",ANALYSIS), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
    } else if (ASSAY == "SCT") {
    PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/",ANALYSIS), pattern = "mm_dream_sct_counts_all.rds", recursive = TRUE, full.names = TRUE)
    }
#Select paths based on all/mild/severe
if (GRADE == "all") { 
    PATHS <- PATHS[!grepl("_severe", PATHS)]
    PATHS <- PATHS[!grepl("_mild", PATHS)]
    print(PATHS)
    } else if (GRADE == "mild") {
    PATHS <- PATHS[grepl("_mild", PATHS)]
    print(PATHS)
    } else if  (GRADE == "severe") {
    PATHS <- PATHS[grepl("_severe", PATHS)]
    print(PATHS)
    } 
RNAlistRDS <- lapply(PATHS, readRDS)
RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE) #combine sub-lists
RNAlistRDS <- RNAlistRDS[!duplicated(RNAlistRDS)]
lapply(RNAlistRDS, gProfiler_FUN, ANALYSIS = ANALYSIS,GRADE = GRADE,ORDERED_QUERY = FALSE, ASSAY = ASSAY, dir_out = dir_out)
lapply(RNAlistRDS, gProfiler_FUN, ANALYSIS = ANALYSIS,GRADE = GRADE,ORDERED_QUERY = TRUE, ASSAY = ASSAY, dir_out = dir_out)
}
